---
title: "732A91 Lab 3"
author: "Fanny Karelius (fanka300), Milda Poceviciute (milpo192)"
date: "3 maj 2018"
output: html_document
---

#Question 1: Normal model, mixture of normal model with semi-conjugate prior

##a) Normal model

The daily precipitation $y=\{y_1,...,y_n\}$ is assumed to be independently normally distributed, $y|\mu,\sigma^2\sim N(\mu,\sigma^2)$, where $\mu,\sigma^2$ are unknown. The priors $\mu\sim N(\mu_0,\tau_0^2)$ and $\sigma^2\sim Inv-\chi^2(\nu_0,\sigma_0^2)$ are independent.
The conditional posteriors are
$$\mu|\sigma^2,y\sim N(\mu_n,\tau_n^2)$$
and
$$\sigma^2|\mu,y\sim Inv-\chi^2(\nu_n,\frac{\nu_0\sigma_0^2+\sum_{i=1}^n(y_i-\mu)^2}{\nu_0+n})$$
where $\mu_n=w\bar{y}+(1-w)\mu_0$, $w=\frac{n/\sigma^2}{\frac{n}{\sigma^2}+\frac{1}{\tau_0^2}}$, $\frac{1}{\tau_n^2}=\frac{n}{\sigma^2}+\frac{1}{\tau_0^2}$ and $\nu_n=\nu_0+n$.

###i)

Gibbs sampling implementation:
```{r}
gibbs_sampler <- function(iter, data, mu0, tau20, nu0, sigma20){
  N <- nrow(data)
  x_bar <- mean(data$V1)
  nun <- nu0+N
  mu <- rnorm(1,mu0, sqrt(tau20))
  sigma2 <- nu0*sigma20/rchisq(1,nu0)
  results <- matrix(ncol=2,nrow=iter+1)
  results[1,1]<-mu
  results[1,2]<-sigma2
  colnames(results)<-c("mu", "sigma2")
  
  for(i in 1:iter){
    w <- (N/results[i,2])/((N/results[i,2])+(1/tau20))
    mun <- w*x_bar+(1-w)*mu0
    tau2n <- 1/((N/results[i,2])+(1/tau20))
    mu <- rnorm(1,mun, sqrt(tau2n))
    para_n <- (nu0*sigma20+sum((data$V1-mu)^2))/nun
    sigma2 <- nun*para_n/rchisq(1,nun) 
    results[i+1,]<-c(mu, sigma2)
  }
  results
}
```


###ii)
```{r}
rain<-read.table("D:/LiU/732A91/Bayesian_lab3/Rainfall.dat",header=FALSE)

mu0<-0
tau20 <-50
nu0 <-5
sigma20 <-20

burn_in <- 100
n <- 1000
rain_gibbs<-gibbs_sampler(n,rain, mu0, tau20, nu0, sigma20)
rain_gibbs <- as.data.frame(rain_gibbs)

post_mu <- mean(rain_gibbs$mu[burn_in:n])
post_sigma2 <- mean(rain_gibbs$sigma2[burn_in:n])
```

```{r,echo=FALSE}
plot(rain_gibbs$sigma2[burn_in:n], xaxt="n",type="l", xlab="Iteration", ylab="sigma2", main ="Gibbs sampling of sigma2")
axis(1, at=seq(0,(n-burn_in), by=50), labels=seq(burn_in,n, by=50))
```

```{r,echo=FALSE}
plot(rain_gibbs$mu[burn_in:n], xaxt="n", type="l", xlab="Iteration", ylab="mu", main ="Gibbs sampling of mu")
axis(1, at=seq(0,(n-burn_in), by=50), labels=seq(burn_in,n, by=50))

```

##b) Mixture normal model

```{r}
rawData <- rain
x <- as.matrix(rawData)

# Model options
nComp <- 2    # Number of mixture components

# Prior options
alpha <- 10*rep(1,nComp) # Dirichlet(alpha)
muPrior <- rep(0,nComp) # Prior mean of mu
tau2Prior <- rep(10,nComp) # Prior std of mu
sigma2_0 <- rep(var(x),nComp) # s20 (best guess of sigma2)
nu0 <- rep(4,nComp) # degrees of freedom for prior on sigma2

# MCMC options
nIter <- 1000 # Number of Gibbs sampling draws

################   END USER INPUT ###############

###### Defining a function that simulates from the 
rScaledInvChi2 <- function(n, df, scale){
  return((df*scale)/rchisq(n,df=df))
}

####### Defining a function that simulates from a Dirichlet distribution
rDirichlet <- function(param){
  nCat <- length(param)
  piDraws <- matrix(NA,nCat,1)
  for (j in 1:nCat){
    piDraws[j] <- rgamma(1,param[j],1)
  }
  piDraws = piDraws/sum(piDraws) # Diving every column of piDraws by the sum of the elements in that column.
  return(piDraws)
}

# Simple function that converts between two different representations of the mixture allocation
S2alloc <- function(S){
  n <- dim(S)[1]
  alloc <- rep(0,n)
  for (i in 1:n){
    alloc[i] <- which(S[i,] == 1)
  }
  return(alloc)
}

# Initial value for the MCMC
nObs <- length(x)
S <- t(rmultinom(nObs, size = 1 , prob = rep(1/nComp,nComp))) # nObs-by-nComp matrix with component allocations.
mu <- quantile(x, probs = seq(0,1,length = nComp))
sigma2 <- rep(var(x),nComp)
probObsInComp <- rep(NA, nComp)

# Setting up the plot
xGrid <- seq(min(x)-1*apply(x,2,sd),max(x)+1*apply(x,2,sd),length = 100)
xGridMin <- min(xGrid)
xGridMax <- max(xGrid)
mixDensMean <- rep(0,length(xGrid))
effIterCount <- 0
ylim <- c(0,2*max(hist(x)$density))


for (k in 1:nIter){
  #message(paste('Iteration number:',k))
  alloc <- S2alloc(S) # Just a function that converts between different representations of the group allocations
  nAlloc <- colSums(S)
  # Update components probabilities
  pi <- rDirichlet(alpha + nAlloc)
  
  # Update mu's
  for (j in 1:nComp){
    precPrior <- 1/tau2Prior[j]
    precData <- nAlloc[j]/sigma2[j]
    precPost <- precPrior + precData
    wPrior <- precPrior/precPost
    muPost <- wPrior*muPrior + (1-wPrior)*mean(x[alloc == j])
    tau2Post <- 1/precPost
    mu[j] <- rnorm(1, mean = muPost, sd = sqrt(tau2Post))
  }
  
  # Update sigma2's
  for (j in 1:nComp){
    sigma2[j] <- rScaledInvChi2(1, df = nu0[j] + nAlloc[j], scale = (nu0[j]*sigma2_0[j] + sum((x[alloc == j] - mu[j])^2))/(nu0[j] + nAlloc[j]))
  }
  
  # Update allocation
  for (i in 1:nObs){
    for (j in 1:nComp){
      probObsInComp[j] <- pi[j]*dnorm(x[i], mean = mu[j], sd = sqrt(sigma2[j]))
    }
    S[i,] <- t(rmultinom(1, size = 1 , prob = probObsInComp/sum(probObsInComp)))
  }
  
}

hist(x, breaks = 20, freq = FALSE, xlim = c(xGridMin,xGridMax), main = "Final fitted density")
lines(xGrid, mixDensMean, type = "l", lwd = 2, lty = 4, col = "red")
lines(xGrid, dnorm(xGrid, mean = post_mu, sd = sqrt(post_sigma2)), type = "l", lwd = 2, col = "blue")
legend("topright", box.lty = 1, legend = c("Data histogram","Mixture density","Normal density from a)"), col=c("black","red","blue"), lwd = 2)
```

##c)


#Question 2: Time series models in Stan

##a)

##b)

##c)

##d)


#Appendix
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```